\bf Mechanistic models for panel data: analysis of ecological experiments with four interacting species

### 2024-04-24 ms.Rnw
Update

<<Setup, include=FALSE,echo=FALSE,results='hide'>>=
 library(plyr)
 library(reshape2)
library(patchwork)
 library(magrittr)
 library(foreach)
 library(readxl)
 library(doParallel)
 registerDoParallel(cores=36)
 library(pomp)
 library(panelPomp)
 library(ggplot2)
 library(gridExtra)
 library(scales)  # For formatting scales
 library(patchwork)

@@

                      if (Sn < 0.0) {
                        Sn = 0.0;
                        error_count += 1;
                      }
                      if (Si < 0.0) {
                        Si = 0.0;
                        error_count += 1000000;
                      }
                      if (F < 0) {
                        F = 0;
                        error_count += 1000;
                      }
                      if (In < 0) {
                        In = 0;
                        error_count += 0.001;
                      }
                      if (Ii < 0) {
                        Ii = 0;
                        error_count += 0.000000001;
                      }
                      if (P < 0 && t > 3.9) {
                        P = 0;
                        error_count += 0.000001;
                      }

                      T_Sn += delta * Sn * dt;
                      T_In += delta * In * dt;
                      T_Si += delta * Si * dt;
                      T_Ii += delta * Ii * dt;

@@

 # Initial state. Assume t0 = day 4
 mix_par_init = Csnippet("
                     Sn = 2.333; //2.3333 = 35/15L
                     Si = 0.667; //0.667 = 10/15L
                     F = 16.667;
                     T_Sn = 0;
                     T_Si = 0;
                     T_In = 0;
                     T_Ii = 0;
                     In = 0;
                     Ii = 0;

@@

                 if (error_count > 0.0) {
                   lik = -100;
                  }
                 else{
                    if(give_log){
                    lik  = dnbinom_mu(lumadult,k_Si,T_Si,give_log) +  dnbinom_mu(luminf,k_Ii,T_Ii,give_log) + dnbinom_mu(dentadult,k_Sn,T_Sn,give_log) +  dnbinom_mu(dentinf,k_In,T_In,give_log);
                    }
                    else{
                    lik  = dnbinom_mu(lumadult,k_Si,T_Si,give_log) *  dnbinom_mu(luminf,k_Ii,T_Ii,give_log) * dnbinom_mu(dentadult,k_Sn,T_Sn,give_log) *  dnbinom_mu(dentinf,k_In,T_In,give_log);
                    }
                      }

@@

\newpage
\spacingset{1.9} % DON'T change the spacing!
\section{Introduction}
\label{sec:intro}
Ecological experiments contain many elements or processes that are difficult to observe, models that can describe both observable and unobservable processes to explain the whole ecological experiment are necessary. As a mathematical description of the elements forming a system, mechanistic modeling can describe the mutual interactions between elements in the system and their interactions with the environment \citep{Stalidzans2020}. Utilizing the established process, they elucidate the actions of hidden mechanisms  The purpose of mechanistic modeling in ecology involves employing mathematical models to depict the development and interaction of dynamic systems. This quantitative tool allows researchers to evaluate the explanatory potential of various conceptual models and measure the magnitude of pertinent parameters. Mechanistic modeling has been widely employed in social sciences, industrial engineering, and physics for an extended period. Improved computational capabilities now make it more effective. Traditional time series analysis offers limited explanatory power due to the nonstationary, nonlinear, and stochastic nature of ecological information. Despite the creation of theoretical models for the data, connecting theories with data continues to be a challenging task. As a result, ecological mechanical modeling cannot solve most of the problems on a wide scale, but rather, only a limited range of the problems \\

Panel data, sometimes referred to as longitudinal data, is composed of multiple time series. Each of these time series, which may be multivariate, contains a series of observations collected from a separate entity  \citep{Carles2020}. In ecology, panel data is a traditional data form that researchers gather to mitigate the impact of random variations throughout experiments, so that it can more accurately measure changes in experimental objectives and control the variables needed to achieve experimental goals. Due to the non-linear characteristics of ecological data, the analysis of panel data entails a high-dimensional structure. Such features diminish the explanatory power of traditional time series analysis and certain Monte Carlo inference techniques \citep{johnstone2009statistical}. The goal of this study is to investigate the practicality of employing mechanistic models on ecological panel data using the panelPOMP structure.\\

The PanelPOMP approach modifies the partially observable Markov Process model to accommodate panel data by constructing a POMP model for each individual component \citep{King2016}. This method enables the determination of whether the parameters within the selected mathematical model are shared or distinct throughout the panel, by computing the highest log likelihood for all potential combinations through the use of panel iterated filtering. Stemming from iterated filtering \citep{King2016} applied to a singular nonlinear stochastic time series, panel iterated filtering not only filters within each series but also cycles across the panel to achieve the best likelihood. This enables the application of mechanistic models to panel data while preserving its intricate dimensionality.\\

@@

For a more in-depth examination of the concealed processes in the panel data, this research incorporates the Susceptible-Infected-Recovered-Food-Parasite (SIRPF) model - within the panelPOMP framework. After applying the model to the \textit{Daphnia} panel data, the maximum log-likelihood and best parameter estimates are determined via an iterative filtering approach. These findings are subsequently compared to discover a more accurate and appropriate model for the given data. The superior performance of the SIRPF model implies that the population density of alga \textit{Ankistrodesmus falcatus}, which is \textit{Daphnia}'s prey, impact the dynamics of both species of \textit{Daphnia}. It is worth noting that the population density of alga is neither observable nor recorded in the experiment. That implies the mechanistic models under panelPOMP potentially better explain the latent process and show the importance of unobservable processes \citep{Edwards2012}. At the same time, the panelPOMP framework provides an estimation of the parameters in the model, which helps to understand the hidden mechanisms in the ecological system \citep{Carles2020}. Based on the estimation, a detailed interpretation of relationship between invasive host and native host with presence of parasite is explaned.\\

Our objective is to evaluate the explanatory capabilities of two traditional ecological models when applied to the \textit{Daphnia} panel data, and to estimate the parameter sets for a more accurate depiction of the hidden mechanisms within the ecological system. By analyzing \textit{Daphnia} panel data, we hope to show the limitations and interpretations of mechanistic models in dealing with ecological data. Additionally, the later section will further discuss how the panelPOMP framework and PIF contribute to overcoming the inherent challenges associated with fitting ecological models.

@@

The dynamics of our study are derived from the Searle's comparison experiment, which focuses on two \textit{Daphnia} species: \textit{Daphnia} \textit{dentifera Forbes}, natively prevalent in North American stratified lakes \citep{searle2016salinization}, and \textit{Daphnia} \textit{lumholtzi Sars}, an invasive species originating from Africa, Asia, and Australia. D. lumholtzi, first observed in North America in the 1990s, has rapidly expanded across the United States, impacting native species and ecosystem dynamics \citep{havel1993daphnia}. These \textit{Daphnia} species, referred to as the native and invasive hosts, are bred in controlled laboratory environments using isofemale lines. For their experiments, six clones of each species were selected to eliminate selection bias \citep{Searle2016}.\\

The dynamics also involves \textit{Metschnikowia bicuspidata}, a parasitic fungus predominantly infecting the native host. This parasite, ingested as spores by \textit{Daphnia}, aggressively affects the host's fecundity and longevity \citep{duffy2010temporal}. The parasite's presence is identifiable by the opaqueness it causes in normally transparent hosts, with spore release upon host death. Searle's study utilized a M. bicuspidata strain from Baker Lake, Michigan, with consistent infectivity and virulence characteristics in different samples \citep{Searle2016}. The final species in our study is \textit{Ankistrodesmus falcatus}, a green alga that serves as the primary food source for \textit{Daphnia}. \textit{Ankistrodesmus falcatus} is a globally distributed freshwater chlorophyte, is known for its role in aquatic communities and has been utilized in ecological research due to its nutritional properties \citep{arnold1971ingestion}. Its significance is particularly evident in controlled experimental settings where the growth conditions and feeding rates of \textit{Daphnia} species can be closely monitored and manipulated \citep{martinez1994effect}.\\

Drawing upon the empirical findings from Searle's laboratory, it was observed that while the two \textit{Daphnia} host species exhibited comparable resource acquisition rates, the invasive species demonstrated a more rapid reproductive cycle \citep{Searle2016}. This observation suggests a potential competitive advantage for the invasive species. Building on these insights, Searle and colleagues structured the ecological system we analyze here, focusing on dynamics at the community level. Their approach involved a mesocosm experiment that varied species composition and parasite presence. The experimental design was a 3x2x1 factorial arrangement, incorporating three species combinations (native only, invasive only, or both existing species), two parasite exposure scenarios (exposed or unexposed), and a single algal food source. This design enabled an in-depth analysis of the invasive host's impact on native populations, manifesting either directly through competition for resources, indirectly via the amplification of a mutual parasite, or through a reduction in parasite effects.\\

In essence, the impact of an invasive \textit{D. lumholtzi}, particularly in terms of resource competition like food, can be significant. This scenario would lead to a decreased survival rate of the native host in the presence of the invasive species, irrespective of parasite interference. This effect is quantifiable through the ratio of natality to mortality in the native host population. Conversely, should the invasive species primarily influence the native host indirectly via interactions with parasites, a different pattern emerges. Here, the survivability of the native host is expected to be lower in scenarios involving both the parasite and the invasive host, as opposed to situations where the native host is only affected by the parasite. This effect is quantifiable through the ratio of natality to mortality in the native host population as well.\\

@@

In the Non-parasitized level, indoor mesocosms were used \citep{Searle2016}, each comprising a 18.9 L bucket with 15 L of high-hardness COMBO media. The experimental design involved 10 replicates for single-species treatments, and 9 replicates for mixed-species treatment. A nitrogen-to-phosphorus ratio of 25:1 was established, and each mesocosm was seeded with 2.5 $\times$ $10^8$ cells of \textit{Ankistrodesmus falcatus}. The laboratory conditions included an average temperature of 23.27$^\circ C$ ($\pm$0.2) and a 16-hour light, 8-hour dark cycle. For single-species treatments, either 45 native or invasive \textit{Daphnia} hosts were added, representing six genotypes per species to eliminate the selection bias. In mixed-species treatments, 35 native and 10 invasive hosts were introduced, simulating early invasion stages.\\

@@

As juveniles rarely exhibit clear signs of infection, only adult \textit{Daphnia} were used for quantifying effects in both non-parasitized and prarasitized dynamics to minimize information bias. Figures below illustrates these dynamics, with the first row of Figure 1 depicting non-parasitized dynamics (mixed-species) where changes in native and invasive host densities reflect direct resource competition, for example, food alga \textit{Ankistrodesmus falcatus}. The second row of Figure 1 shows parasitized dynamics, where the interplay of susceptible and infected \textit{Daphnia} of different species is further influenced by parasites, complicating the variation of susceptible and infected \textit{Daphnia} of different races, which makes it difficult to directly analyze the sources of influence on their variation. Non-parasitized dynamics will be analyzed at the first place to help understand the effect of resource competition in mixed-species. And based on this result, the correlation between native host and invasive host in parasitized dynamics will be further explored.
% 
\begin{figure}[H]
% \centering
<<echo=FALSE, fig.width=8, fig.height=6,out.width="2in">>=
 plot(x = 1:10)
@
\end{figure}
%

\begin{figure}[H]
\centering
<<DataVis,echo=FALSE, fig.width=9,fig.height=6,out.width="2in">>=
mixed_nopara_data_combined <- do.call(rbind, lapply(1:length(mixed_nopara_data), function(i) {
  cbind(mixed_nopara_data[[i]], Trial = paste("Trial", trails[i]))
}))
mixed_nopara_data_combined$Bucket = paste0(rep("Bucket ",80),rep(1:9, each = 10))

@@

# Adjusting p1 with specified font size modifications
p1 <- ggplot(mixed_nopara_data_combined, aes(x = day, y = sqrt(dent.adult))) +
  geom_line(aes(color = "Dent Adult"), size = 1) +
  geom_line(aes(y = sqrt(lum.adult), color = "Lum Adult"), size = 1) +
  facet_wrap(~Bucket, nrow = 1, labeller = bucket_labeller) +
  theme(axis.title.x = element_blank( ),
        axis.title.y = element_text(size = 10),  # Y-axis title
        axis.text.x = element_text(size = 10),   # X-axis text
        axis.text.y = element_text(size = 10),   # Y-axis text
        legend.title = element_text(size = 10),  # Legend title
        legend.text = element_text(size = 10),   # Legend text
        strip.text = element_text(size = 10),    # Facet label text
        strip.background = element_blank(),
        strip.placement = "outside",
        plot.margin = c(.5,.5,.5,.5)) +
  ylab('Square root of Daphnia density') +
  scale_x_continuous(limits = c(0, 52), breaks = c(0, 25, 50)) +
  labs(title = "Daphnia Density in Non-parasitized Dynamics", color = "Species") +
  scale_color_manual(values = c("Dent Adult" = "black", "Lum Adult" = "blue")) +
  theme_minimal() +
  guides(color = guide_legend(title = "Species"))

# Adjusting p2 with specified font size modifications
p2 <- ggplot(mixed_para_data_combined, aes(x = day, y = sqrt(dentadult))) +
  geom_line(aes(color = "Dent Adult"), size = 1) +
  geom_line(aes(y = sqrt(lumadult), color = "Lum Adult"), size = 1) +
  geom_line(aes(y = sqrt(luminf), color = "Lum Inf"), size = 1) +
  geom_line(aes(y = sqrt(dentinf), color = "Dent Inf"), size = 1) +
  facet_wrap(~Bucket, nrow = 1, labeller = bucket_labeller) +
  theme(axis.title.x = element_text(size = 10),
        axis.title.y = element_text(size = 10),
        axis.text.x = element_text(size = 10),
        axis.text.y = element_text(size = 10),
        legend.title = element_text(size = 10),
        legend.text = element_text(size = 10),
        strip.text = element_text(size = 10),
        strip.background = element_blank(),
        strip.placement = "outside",
        plot.margin = c(.5,.5,.5,.5)) +
  ylab('Square root of Daphnia density') +
  scale_x_continuous(limits = c(0, 52), breaks = c(0, 25, 50)) +
  labs(title = "Daphnia Density in Parasitized Dynamics", color = "Species") +
  scale_color_manual(values = c("Dent Adult" = "black", "Lum Adult" = "blue", "Lum Inf" = "green", "Dent Inf" = "red")) +
  theme_minimal() +
  guides(color = guide_legend(title = "Species"))

@@

@
% \includegraphics[width=1\textwidth]{"Density_plot.png"}
\label{fig:plot1}
\caption{Plot of density of hosts in dynamics including both native and invasive host species of \textit{Daphnia}. The Buckets 1 to 9 represents different buckets in different experiment. The black lines represent the \textit{D. dentifera} density, the red lines show the \textit{D. lumholtzi} density, the green lines are the infected \textit{D. dentifera} density, and the blue lines indicate the infected \textit{D. lumholtzi} density.}
\end{figure}

@@

Methodologically, POMP models are analyzed using a suite of computational techniques. Sequential Monte Carlo methods, such as particle filters, are commonly employed for state estimation and parameter inference in these models \citep{bhadra2011malaria}. These methods involve simulating a large number of potential realizations of the latent process and updating these simulations as new observations become available, thereby approximating the posterior distribution of the state given the data \citep{King2016}. Additionally, techniques like iterated filtering and particle Markov chain Monte Carlo are used for parameter estimation and model calibration.\\

@@

Working with POMP models can prove difficult due to the complexity of formulating both the likelihood function and transition density in intricate settings. To address this issue, several techniques have been devised, including the particle filter. Utilizing the particle filter method, it is only necessary to supply data, initial density simulators, single time-step transition density simulators for the latent process, and an evaluator for the observation process density reliant on the hidden process in order to achieve maximum likelihood estimates.\\

To implement the particle filter, we start by initializing $M$ particles at time 1, where $\{ X_1^m; m \in 1:M \}$. All with the required latent state detail and a vector of parameter values, $\theta$. As a result, for every time step n, where $n \in $ [$N$], we advance the each particle by doing the one-step random walk based on transition density, resulting in a set of particles that represents the prediction distribution at time n, which can be represented as $f_{X_n|X_{n-1}}(\cdot | X_{n-1}^m; \theta)$. Next, we assign weights to the particles based on the measurement density to obtain an ensemble that represents the filtering distribution at time n, where the new weight is calculated via $w_{n,m} = f_{Y_n|X_n}(y_n | x_n^m)$. Finally, we reassign the particles based on their weights to generate a new set of particles that represents the filtering distribution, which can then be used to estimate model parameters swarm based on the maximum likelihood \citep{5}. The particle filter is particularly useful in situations where both the likelihood function and transition density are difficult to calculate analytically.\\

@@

\end{tikzpicture}
}
\end{center}
\vspace{-5mm}
\caption{A flow diagram for the SIRPF2 model. Apart from the quantified states above, the graph includes $R$ state, which represents death status. The flow diagram shows the how two species of \textit{Daphnia} interact with each other indirectly through $F$ and $P$. The marginal processes have been shown in SIRPF mode, and SIRPF2 model is the adaption form in mixed-species dynamics}
\label{fig:flow_SIRPF2}
\end{figure}

@@

The integration of these stochastic terms within the SIRPF2 framework, as detailed in Table 1, enhances the model's fidelity in simulating ecological interactions. The model, grounded in the Lotka-Volterra competition framework and augmented with stochastic components, provides a nuanced understanding of the interplay between native and invasive hosts, and their interaction with the surrounding resources and parasites. This incorporation of stochasticity, particularly through the lens of Brownian motion, is pivotal in transcending the limitations of deterministic models, thereby offering a more robust and comprehensive tool for ecological analysis.\\

@@

@
When constructing statistical models to analyze ecological data, the specification of parameters plays a crucial role in shaping the model's complexity and interpretability. In a scenario with 10 units of observation, the decision to assign a parameter as shared across units or unique to each one has significant implications for the model's dimensionality. For instance, designating a parameter as shared across all units incrementally increases the model dimension by one, whereas categorizing a parameter as unique to each unit multiplies the dimensionality by the number of units, which in this case results in an increase by 10 for single species Non-parasitized dynamics, 9 for mixed species Non-parasitized dynamics and Parasitized dynamics with the invasive species of \textit{Daphnia}, and 8 for the rest dynamics.\\

>>>>

<<Setup, include=FALSE,echo=FALSE,results='hide'>>=
 library(plyr)
 library(reshape2)
library(patchwork)
 library(magrittr)
 library(foreach)
 library(readxl)
 library(doParallel)
 registerDoParallel(cores=36)
 library(pomp)
 library(panelPomp)
 library(ggplot2)
 library(gridExtra)
 library(scales)  # For formatting scales
 library(patchwork)
 library(dplyr)
 library(gtable)

@@

                      if (Sn < 0.0 || Sn > 1e5) {
                        Sn = 0.0;
                        error_count += 1;
                      }
                      if (Si < 0.0 || Si > 1e5) {
                        Si = 0.0;
                        error_count += 1000000;
                      }
                      if (F < 0.0 || F > 1e20) {
                        F = 0.0;
                        error_count += 1000;
                      }
                      if (In < 0.0 || In > 1e5) {
                        In = 0.0;
                        error_count += 0.001;
                      }
                      if (Ii < 0.0 || Ii > 1e5) {
                        Ii = 0.0;
                        error_count += 0.000000001;
                      }                      
                      if ((P < 0.0 || P > 1e20)&& t > 3.9) {
                        P = 0.0;
                        error_count += 0.000001;
                      }

                      T_Sn = delta * Sn * dt;
                      T_In = delta * In * dt;
                      T_Si = delta * Si * dt;
                      T_Ii = delta * Ii * dt;

                      T_Sn = fabs(T_Sn);
                      T_In = fabs(T_In);
                      T_Si = fabs(T_Si);
                      T_Ii = fabs(T_Ii);

@@

# Initial state. Assume t0 = day 4
mix_par_init = Csnippet("
                     Sn = 2.333; //2.3333 = 35/15L
                     Si = 0.667; //0.667 = 10/15L
                     F = 16.667;

                     T_Sn = 0.0;
                     T_Si = 0.0;
                     T_In = 0.0;
                     T_Ii = 0.0;
                     In = 0.0;
                     Ii = 0.0;

@@

                 if (error_count > 0.0) {
                   lik = -150;
                  }
                 else{
                    if(give_log){
                    lik  = dnbinom_mu(lumadult,k_Si,T_Si,give_log) +  dnbinom_mu(luminf,k_Ii,T_Ii,give_log) + dnbinom_mu(dentadult,k_Sn,T_Sn,give_log) +  dnbinom_mu(dentinf,k_In,T_In,give_log);
                    }
                    else{
                    lik  = dnbinom_mu(lumadult,k_Si,T_Si,give_log) *  dnbinom_mu(luminf,k_Ii,T_Ii,give_log) * dnbinom_mu(dentadult,k_Sn,T_Sn,give_log) *  dnbinom_mu(dentinf,k_In,T_In,give_log);
                    }
                      }

@@

\newpage
\spacingset{1.9} % DON'T change the spacing!
\section{Introduction}
\label{sec:intro}
Ecological experiments contain many elements or processes that are difficult to observe, models that can describe both observable and unobservable processes to explain the whole ecological experiment are necessary. As a mathematical description of the elements forming a system, mechanistic modeling can describe the interactions between elements in the dynamic system and their interactions with the environment \citep{Stalidzans2020}. Utilizing the established process by mechanistic models, we can understand the hidden mechanisms in dynamic systems better. This quantitative tool allows researchers to evaluate the explanatory potential of various conceptual models and measure the magnitude of pertinent parameters. Mechanistic modeling has been widely employed in social sciences, industrial engineering, and physics for an extended period. Improved computational capabilities now make it more effective. The purpose of mechanistic modeling in ecology involves employing mathematical models to depict the development and interaction of dynamic systems. Traditional time series analysis offers limited explanatory power due to the nonstationary, nonlinear, and stochastic nature of ecological information. Despite the creation of theoretical models for the data, connecting theories with data continues to be a challenging task. As a result, ecological mechanical modeling cannot solve most of the problems on a wide scale, but rather, only a limited range of the problems. \\

Panel data, sometimes referred to as longitudinal data, is composed of multiple time series. Each of these time series, which may be multivariate, contains a series of observations collected from a separate entity \citep{Carles2020}. In ecology, panel data is a traditional data form that researchers gather to mitigate the impact of random variations throughout experiments, so that it can more accurately measure changes in experimental objectives and control the variables needed to achieve experimental goals. The PanelPOMP approach modifies the partially observable Markov Process model to accommodate panel data by constructing a POMP model for each individual component \citep{King2016}. This method enables the determination of whether the parameters within the selected mathematical model are shared or distinct throughout the panel, by computing the highest log likelihood for all potential combinations through the use of panel iterated filtering. Stemming from iterated filtering \citep{ionides2011iterated} applied to a singular nonlinear stochastic time series, panel iterated filtering not only filters within each series but also cycles across the panel to achieve the best likelihood. This enables the application of mechanistic models to panel data while preserving its intricate dimensionality.\\

Due to the non-linear characteristics of ecological data, the analysis of panel data entails a high-dimensional structure. Such features diminish the explanatory power of traditional time series analysis and certain Monte Carlo inference techniques \citep{johnstone2009statistical}. The goal of this study is to investigate the practicality of employing mechanistic models on ecological panel data using the panelPOMP structure.\\

@@

For a more in-depth examination of the concealed processes in the panel data, this research incorporates the Susceptible-Infected-Recovered-Food-Parasite (SIRPF) model - within the panelPOMP framework. After applying the model to the \textit{Daphnia} panel data, the best parameter estimates are determined through maximum log-likelihood estimation via the iterative filtering approach. These findings are subsequently compared to discover a more accurate and appropriate model for the given data. The good performance of the SIRPF model implies that the population density of alga \textit{Ankistrodesmus falcatus}, which is \textit{Daphnia}'s prey, plays an important role in the dynamics of both species of \textit{Daphnia}. It is worth noting that the population density of alga is neither observable nor recorded in the experiment. That implies the mechanistic models under panelPOMP potentially better explain the latent process and show the importance of unobservable processes \citep{Edwards2012}. Nevertheless, the panelPOMP framework provides an estimation of the parameters for each dynamic systems included, which helps to understand the hidden mechanisms in the ecological system \citep{Carles2020}. Based on the estimation, a detailed interpretation of relationship between each component in the ecological system is explained.\\

Our objective is to evaluate the explanatory capabilities of SIRPF model when applied to the \textit{Daphnia} panel data, and to estimate the parameter sets for a more accurate interpretation of the hidden mechanisms within the ecological system comparing to classical model. By analyzing \textit{Daphnia} panel data, we also show the interpretations and limitations of mechanistic models when dealing with ecological data. Additionally, how the panelPOMP framework and PIF contribute to overcoming the inherent challenges associated with fitting ecological models will be discussed in detail.

@@

The dynamics of our study are derived from the Searle's comparison experiment, which focuses on two \textit{Daphnia} species: \textit{Daphnia} \textit{dentifera Forbes}, natively prevalent in North American stratified lakes \citep{searle2016salinization}, and \textit{Daphnia} \textit{lumholtzi Sars}, an invasive species originating from Africa, Asia, and Australia. \textit{D. lumholtzi}, first observed in North America in the 1990s, has rapidly expanded across the United States, impacting native species and ecosystem dynamics \citep{havel1993daphnia}. These \textit{Daphnia} species, referred to as the native and invasive hosts, are bred in controlled laboratory environments using isofemale lines. For their experiments, six clones of each species were selected to eliminate selection bias \citep{Searle2016}.\\

The dynamics also involves \textit{Metschnikowia bicuspidata}, a parasitic fungus predominantly infecting the native host. This species of parasite is strain from Baker Lake, Michigan, with consistent infectivity and virulence characteristics in different samples \citep{Searle2016}. Searle's study utilized the \textit{M. bicuspidata}, which is ingested as spores by \textit{Daphnia}, aggressively affects the host's fecundity and longevity \citep{duffy2010temporal}. The parasite's presence is identifiable by the opaqueness it causes in normally transparent hosts, with spore release upon host death.  \textit{Ankistrodesmus falcatus}, a green alga that serves as the primary food source for \textit{Daphnia} also presents in the experiment. \textit{Ankistrodesmus falcatus} is a globally distributed freshwater chlorophyte, is known for its role in aquatic communities and has been utilized in ecological research due to its nutritional properties \citep{arnold1971ingestion}. Its significance is particularly evident in controlled experimental settings where the growth conditions and feeding rates of \textit{Daphnia} species can be closely monitored and manipulated \citep{martinez1994effect}.\\

Drawing upon the empirical findings from Searle's laboratory, it was observed that while the two \textit{Daphnia} host species exhibited comparable resource acquisition rates, the invasive species is a strong interspecific competitor \citep{Searle2016}. This observation suggests a potential competitive advantage for the invasive species. Building on these insights, Searle and colleagues structured the ecological system we analyze here, focusing on dynamics at the community level. Their approach involved a mesocosm experiment that varied species composition and parasite presence. The experimental design was a 3 $\times$ 2 $\times$ 1 factorial arrangement, incorporating three species combinations (native only, invasive only, or both existing species), two parasite exposure scenarios (exposed or unexposed), and a single algal food source. This design enabled an in-depth analysis of the invasive host's impact on native populations, manifesting either directly through competition for resources, indirectly via the amplification of a mutual parasite, or through a reduction in parasite effects.\\

In essence, the impact of an invasive \textit{D. lumholtzi}, particularly in terms of resource competition like food, can be significant. This scenario would lead to a decreased survival rate of the native host in the presence of the invasive species, irrespective of parasite interference. This effect is quantifiable through the ratio of natality to mortality in the native host population. Apart from food, disease affect species' survivability. The invasive host dilutes disease when it is a strong interspecific competitor, while a weak interspecific invasive species amplifies disease. This effect is quantifiable through the ratio of natality to mortality in the native host population as well.\\

@@

At the non-parasitized level, indoor mesocosms were used \citep{Searle2016}, each comprising a 18.9 L bucket with 15 L of high-hardness COMBO media. The experimental design involved 10 replicates for single-species treatments, and 9 replicates for mixed-species treatment. A nitrogen-to-phosphorus ratio of 25:1 was established, and each mesocosm was seeded with 2.5 $\times$ $10^8$ cells of \textit{Ankistrodesmus falcatus}. The laboratory conditions included an average temperature of 23.27$^\circ C$ ($\pm$0.2) and a 16-hour light, 8-hour dark cycle. For single-species treatments, either 45 native or invasive \textit{Daphnia} hosts were added, representing six genotypes per species to eliminate the selection bias. In mixed-species treatments, 35 native and 10 invasive hosts were introduced, simulating early invasion stages.\\

@@

As juveniles rarely exhibit clear signs of infection, only adult \textit{Daphnia} were used for quantifying effects in both non-parasitized and prarasitized dynamics to minimize information bias. Figure 1 illustrates these dynamics, with the first row of Figure 1 depicting non-parasitized dynamics (mixed-species) where changes in native and invasive host densities reflect direct resource competition, for example, food alga \textit{Ankistrodesmus falcatus}. The second row of Figure 1 shows parasitized dynamics, where the interplay of susceptible and infected \textit{Daphnia} of different species is further influenced by parasites, complicating the variation of susceptible and infected \textit{Daphnia} of different races, which makes it difficult to directly analyze the sources of influence on their variation. Non-parasitized dynamics will be analyzed at the first place to help understand the effect of resource competition in mixed-species. And based on this result, the correlation between native host and invasive host in parasitized dynamics will be further explored.

\begin{figure}[H]
\centering
<<DataVis,echo=FALSE, fig.width=9,fig.height=6,out.width="7in">>=
mixed_nopara_data_combined <- do.call(rbind, lapply(1:length(mixed_nopara_data), function(i) {
  cbind(mixed_nopara_data[[i]], Trial = paste("Trial", trails[i]))
}))
mixed_nopara_data_combined$Bucket = paste0(rep("Bucket ",80),rep(1:9, each = 10))

@@

# Adjusting p1 with specified font size modifications
p1 <- ggplot(mixed_nopara_data_combined, aes(x = day, y = sqrt(dent.adult))) +
  geom_line(aes(color = "Dent Susceptible"), linewidth = 1) +
  geom_line(aes(y = sqrt(lum.adult), color = "Lum Susceptible"), linewidth = 1) +
  facet_wrap(~Bucket, nrow = 1, labeller = bucket_labeller) +
  theme(axis.title.x = element_blank( ),
        axis.title.y = element_text(size = 10),  # Y-axis title
        axis.text.x = element_text(size = 10),   # X-axis text
        axis.text.y = element_text(size = 10),   # Y-axis text
        legend.title = element_text(size = 10),  # Legend title
        legend.text = element_text(size = 10),   # Legend text
        strip.text = element_text(size = 10),    # Facet label text
        strip.background = element_blank(),
        strip.placement = "outside",
        plot.margin = c(.5,.5,.5,.5)) +
  ylab('Square root of Daphnia density per liter') +
  scale_x_continuous(limits = c(0, 52), breaks = c(0, 25, 50)) +
  labs(title = "Daphnia Density in Non-parasitized Dynamics", color = "Species") +
  scale_color_manual(values = c("Dent Susceptible" = "black", "Lum Susceptible" = "blue")) +
  theme_minimal() +
  guides(color = guide_legend(title = "Species"))

# Adjusting p2 with specified font size modifications
p2 <- ggplot(mixed_para_data_combined, aes(x = day, y = sqrt(dentadult))) +
  geom_line(aes(color = "Dent Susceptible"), linewidth = 1) +
  geom_line(aes(y = sqrt(lumadult), color = "Lum Susceptible"), linewidth = 1) +
  geom_line(aes(y = sqrt(luminf), color = "Lum Inf"), linewidth = 1) +
  geom_line(aes(y = sqrt(dentinf), color = "Dent Inf"), linewidth = 1) +
  facet_wrap(~Bucket, nrow = 1, labeller = bucket_labeller) +
  theme(axis.title.x = element_text(size = 10),
        axis.title.y = element_text(size = 10),
        axis.text.x = element_text(size = 10),
        axis.text.y = element_text(size = 10),
        legend.title = element_text(size = 10),
        legend.text = element_text(size = 10),
        strip.text = element_text(size = 10),
        strip.background = element_blank(),
        strip.placement = "outside",
        plot.margin = c(.5,.5,.5,.5)) +
  ylab('Square root of Daphnia density per liter') +
  scale_x_continuous(limits = c(0, 52), breaks = c(0, 25, 50)) +
  labs(title = "Daphnia Density in Parasitized Dynamics", color = "Species") +
  scale_color_manual(values = c("Dent Susceptible" = "black", "Lum Susceptible" = "blue", "Lum Inf" = "green", "Dent Inf" = "red")) +
  theme_minimal() +
  guides(color = guide_legend(title = "Species"))

@@

@
% \includegraphics[width=1\textwidth]{"Density_plot.png"}
\label{fig:data_vis}
\caption{Plot of density of hosts in dynamics including both native and invasive host species of \textit{Daphnia}. The Buckets 1 to 9 represents different buckets in different experiment. The black lines represent the \textit{D. dentifera} density, the red lines show the \textit{D. lumholtzi} density, the green lines are the infected \textit{D. dentifera} density, and the blue lines indicate the infected \textit{D. lumholtzi} density.}
\end{figure}

@@

Methodologically, POMP models are analyzed using combination of various computational techniques. Sequential Monte Carlo methods, such as particle filters, are commonly employed for state estimation and parameter inference in these models \citep{bhadra2011malaria}. These methods involve simulating a large number of potential realizations of the latent process and updating these simulations as new observations become available, thereby approximating the posterior distribution of the state given the data \citep{King2016}. Additionally, techniques like iterated filtering and particle Markov chain Monte Carlo are used for parameter estimation and model calibration.\\

@@

Working with POMP models can prove difficulty due to the complexity of formulating both the likelihood function and transition density in intricate settings. To address this issue, several techniques have been devised, including the particle filter. Utilizing the particle filter method, it is only necessary to supply data, initial density simulators, single time-step transition density simulators for the latent process, and an evaluator for the observation process density reliant on the hidden process in order to achieve maximum likelihood estimates.\\

To implement the particle filter, we start by initializing $M$ particles at time 1, where $\{ X_1^m; m \in (1,2,\cdots,M) \}$. All with the required latent state detail and a vector of parameter values, $\theta$. As a result, for every time step n, where $n \in $ [$1:N$], we advance the each particle by doing the one-step random walk based on transition density, resulting in a set of particles that represents the prediction distribution at time n, which can be represented as $f_{X_n|X_{n-1}}(\cdot | X_{n-1}^m; \theta)$. Next, we assign weights to the particles based on the measurement density to obtain an ensemble that represents the filtering distribution at time n, where the new weight is calculated via $w_{n,m} = f_{Y_n|X_n}(y_n | x_n^m)$. Finally, we reassign the particles based on their weights to generate a new set of particles that represents the filtering distribution, which can then be used to estimate model parameters swarm based on the maximum likelihood \citep{Carles2020}. The particle filter is particularly useful in situations where both the likelihood function and transition density are difficult to calculate analytically.\\

@@

\end{tikzpicture}
}
\end{center}
\vspace{-5mm}
\caption{A flow diagram for the SIRPF2 model. The graph includes $R$ state, which represents death status. The flow diagram shows the how two species of \textit{Daphnia} interact with each other indirectly through $F$ and $P$. The marginal processes have been shown in SIRPF mode, and SIRPF2 model is the adaption form in mixed-species dynamics}
\label{fig:flow_SIRPF2}
\end{figure}

@@

The integration of these stochastic terms within the SIRPF2 framework, as detailed in Table 1, enhances the model's fidelity in simulating ecological interactions. The model, grounded in the Lotka-Volterra competition framework and augmented with stochastic components, provides a detailed understanding of the interplay between native and invasive hosts, and their interaction with the surrounding resources and parasites. This incorporation of Brownian motion, is pivotal in transcending the limitations of deterministic models, thereby offering a more robust and comprehensive tool for ecological analysis.\\

@@

@
When constructing statistical models to analyze ecological data, the specification of parameters plays a crucial role in shaping the model's complexity and interpretability. In a scenario with 10 units of observation, the decision to assign a parameter as shared across units or unique to each one has strong implications for the model's dimension. For instance, designating a parameter as shared across all units incrementally increases the model dimension by one, whereas categorizing a parameter as unique to each unit multiplies the dimension by the number of units, which in this case results in an increase by 10 for single species non-parasitized dynamics, 9 for mixed species non-parasitized dynamics and parasitized dynamics with the invasive species of \textit{Daphnia}, and 8 for the rest dynamics.\\
